Investigating social change during cultural contact period using geometric morphometry of pottery shapes from Iron Age northeastern Taiwan

### 2019-12-31 analysis/paper/paper.Rmd
some updates, work on introduction a little

---
title: "Investigating social change during cultural contact period using geometric morphometry of pottery shapes from Iron Age northeastern Taiwan"
author:
  - Li-Ying Wang
  - Ben Marwick
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
---

@@

Ceramic specialization can reflect prehistoric socioeconomic patterns since it relates to not only economic base for households but also societies as a whole when specialized production was organized. Identifying the specialization in ceramic production will help to understand how complex societies form and further explore the underlying mechanism. One way to examine ceramic specialization is to identify whether ceramics present standardization in production, which based on the assumption that the specialized mass production will lead to uniformity of the product due to increased skills, routinization, and lower diversity of producers that enable us to interpret social organization or social relations such as type of production, and the presence of elite control [@Arnold2000, pp. 334; @Stark1995; @Costin1991].

@@

Pottery analyzed in this paper comes from Kiwulan, northeastern Taiwan. Kiwulan is situated on a hill near a riverside at the northern margin of Yilan, which is characterized by a triangular alluvial plain facing eastwards the Pacific and mountains on three other sides. Kiwulan can be divided into two cultural components, the upper and the lower component, with a sterile layer in between [@Chen2007]. This paper focus on the upper component dated to AD 1350 to AD 1950, which covers around 600 years from the late Iron Age to the historical period started with the European presence in Taiwan in the early 17th century. The Dutch first came and occupied southern Taiwan in 1624 and then the Spanish occupied northern Taiwan in 1626. In 1642, the Spanish was expelled by the Dutch, who took over their forts in northern Taiwan. Since then, western Taiwan was mostly under the colonial rule of the Dutch until 1662 when the Kingdom of Tungning was founded by Koxinga, a loyalist of the Ming dynasty of China [@Andrade2007].

@@

More than 550 thousands of potsherds were excavated from the upper component at Kiwulan, and around 1,200 vessels were reconstructed. Those vessels show high consistency in shape - globular body with short neck and wide mouth. A wide variety of impressed geometric motifs applied all over the exterior body surface are observed. They were believed as cooking pots according to the evidence of chars commonly observed on both interior and exterior surfaces. Most pots are fired to orange to brownish color with a fully oxidized core or a reduced core with oxidized fringes.

>>>>

---
title: "Investigating social change during cultural contact period using geometric morphometry of pottery shapes from Iron Age northeastern Taiwan"
author:
  - Li-Ying Wang
  - Ben Marwick
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  The emergence of ceramic specialization in prehistoric societies is often linked to shifts in the complexity of social structures, because specialized pottery production can reflect craft specialization and the presence of elite control. Previous work on identifying specialization relies on typological or linear metric analysis. Here we demonstrate how to investigate ceramic standardization by analyzing outlines of ceramic vessels. Outline analysis is an important method because, unlike more commonly-used landmark analysis methods, it can effectively quantify shape differences for objects that lack distinctive measurement points needed for landmark analysis. We demonstrate this method using pottery from Kiwulan, a large multi-component Iron Age site (AD 1350-1850) in northeast Taiwan. To measure ceramic specialization, we quantified pottery standardization by analyzing shape variables with reproducible geometric morphometric methods. We computed coefficients of variation (CVs) for shape coefficients obtained by elliptical Fourier analysis to test for shape standardization. We found significant differences in pottery shape and shape standardization that indicate changes in pottery production resulting from contact with mainland Han Chinese groups in northeast Taiwan. We infer increasing craft specialization and changes in social organization. Our case study, which includes an openly available research compendium of R code, represents an innovative application of outline-based methods in geometric morphometry to answer the anthropological questions of craft specialization. This study implies that craft specialization can be a deliberate act of resistance to show indigenous identity and ethnicity in a direct culture contact situation.
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
---

@@

- the definition of craft specialization

The emergence of craft specialization is usually associated with the changes in a socio-economic system of a society since it links to the organization of production

context, concentration, scale, intensity

Ceramic specialization can reflect prehistoric socio-economic patterns since it relates to not only economic base for households but also societies as a whole when specialized production was organized.

Identifying the specialization in ceramic production will help to understand how complex societies form and further explore the underlying mechanism. One way to examine ceramic specialization is to identify whether ceramics present standardization in production, which based on the assumption that the specialized mass production will lead to uniformity of the product due to increased skills, routinization, and lower diversity of producers that enable us to interpret social organization or social relations such as type of production, and the presence of elite control [@Arnold2000, pp. 334; @Stark1995; @Costin1991].

social changes -  culture contact framework 
- the cause of emergence of craft specialization, elite control, rice intensification, political, economic 
- here I have the third cultural resistance, identity take a look at social boundary
- standardization as indicator, metric and technological
- GM is more sensitive for subtle changes, and the outline
- innovative: CV test 
- hypothesis, goals

Ceramic specialization can reflect prehistoric socio-economic patterns since it relates to not only economic base for households but also societies as a whole when specialized production was organized. Identifying the specialization in ceramic production will help to understand how complex societies form and further explore the underlying mechanism. One way to examine ceramic specialization is to identify whether ceramics present standardization in production, which based on the assumption that the specialized mass production will lead to uniformity of the product due to increased skills, routinization, and lower diversity of producers that enable us to interpret social organization or social relations such as type of production, and the presence of elite control [@Arnold2000, pp. 334; @Stark1995; @Costin1991].

@@

Ceramics analyzed in this paper come from 40 units (4m by 4m) sampled from 4 connected trenches at Kiwulan, northeastern Taiwan. Kiwulan is situated on a hill near a riverside at the northern margin of Yilan, which is characterized by a triangular alluvial plain facing eastwards the Pacific and mountains on three other sides. The chronology of Kiwulan can be divided into two cultural components, the upper and the lower component, with a sterile layer in between [@Chen2007]. This paper focuses on the upper component dating from AD 1350 to AD 1950, which covers the late Iron Age and the historical period defined by the European presence in Taiwan in the early 17th century. The Dutch firstly came and occupied southern Taiwan in 1624 and then the Spanish occupied northern Taiwan in 1626. In 1642, the Spanish was expelled by the Dutch, who took over their forts in northern Taiwan. Since then, western Taiwan was mostly under the colonial rule of the Dutch until 1662 when the Kingdom of Tungning was founded by Koxinga, a loyalist of the Ming dynasty of China [@Andrade2007].

@@

More than 550 thousands of sherds were excavated from the upper component at Kiwulan, and around 1,200 vessels were reconstructed. There are only two forms of vessels, one is cooking pot and another one is steamer made of two pots stacking together. The pots show high consistency in shape - globular body with short neck and wide mouth. A wide variety of impressed geometric motifs are applied all over the exterior body surface. They were believed used for cooking according to the evidence of charred residue commonly observed on interior surface and soot on exterior surface. Besides, there is no other utilitarian vessels found at this site. Most pots are fired to orange to brownish color with a fully oxidized core or a reduced core with oxidized fringes. Those pots might be pinched and shaped by hand according to the pressed finger marks usually observed inside of pots.
